Fully structured proof theory for intuitionistic modal logics

### 2020-06-02 BicolorNS/aiml20/aiml20.tex
changing the repo structure
intro done

%\usepackage[matrix,arrow]{xy}
\usepackage[noxy]{virginialake}
%\vlnosmallleftlabels
\vlnostructuressyntax

@@

\begin{frontmatter}
  \title{Fully structured proof theory \\for intuitionistic modal logics \\(working title)}
  \author{Sonia Marin}
  %\footnote{You can put your email address or grant acknowledgement as a footnote, if you wish.}
  \address{University College London, UK}
  \author{Marianela Morales}
  \address{LIX, \'Ecole Polytechnique  \&  Inria Saclay, France}

@@

\todo{sell the systems! more emphasis in the nested sequent system?}

Structural proof theoretic accounts of intuitionistic modal logic can adopt the paradigm of \emph{labelled deduction} in the form of labelled natural deduction and labelled sequent systems~\cite{simpson1994}, or the one of \emph{unlabelled deduction} in the form of sequent~\cite{bierman2000} or nested sequent systems~\cite{strassburger2013} (for a survey see~\cite[Chap.~3]{marin2018}). In this article, we present a labelled sequent system and a nested sequent system for intuitionistic modal logics capturing the accesibility relation for modal logics and the intuitionistic relation $\le$. In the labelled calculus we made explicity both relations with $\rel$ and $\le$ into the system. In the nested sequent system, we represent this relations with one bracket in a sequent (for the accesibility relation) and a with two brackets in a sequent (preorder relation).

\section{Intuitionistic modal logics}
The language of {intuitionisitic modal logic} is the one of intuitionistic propositional logic with the modal operators $\BOX$ and $\DIA$, standing most generally for \emph{necessity} and \emph{possibility}.
%
Starting with a set $\mathcal{A}$ of atomic propositions, denoted $\fm a$, modal formulas are constructed from the grammar:
%
$$
\fm A \coloncolonequals
\fm a \mid \fm{A \AND A} \mid \fm{A \OR A} \mid \fm{\TOP} \mid \fm{\BOT} \mid \fm{A \IMP A} \mid \fm{\BOX A} \mid \fm{\DIA A}
$$
%
The axiomatisation that is now generally accepted as intuitionistic modal logic $\IK$ was given by Plotkin and Stirling~\cite{plotkin1986} and is equivalent to the one proposed by Fischer-Servi~\cite{fischer1984}.
%, and by Ewald~\cite{Ewald} in the case of intuitionistic tense logic. 
%
It is obtained from intuitionistic propositional logic by adding:
\begin{itemize}
	\item the \emph{necessitation rule}: $\fm{\BOX A}$ is a theorem if $\fm{A}$ is a theorem; and
	\item the following five variants of the \emph{distributivity axiom}:
	\begin{equation*}
	\label{eq:ik}%\hskip-2em
	\begin{array}[t]{r@{\;}l@{\quad}r@{\;}l@{\quad}r@{\;}l}
	\kax[1]\colon&\fm{\BOX(A\IMP B)\IMP(\BOX A\IMP\BOX B)}
	&
	\kax[3]\colon&\fm{\DIA(A\OR B)\IMP(\DIA A\OR\DIA B)}
	&
	\kax[5]\colon&\fm{\DIA\BOT\IMP\BOT}
	\\
	\kax[2]\colon&\fm{\BOX(A\IMP B)\IMP(\DIA A\IMP\DIA B)}
	&
	\kax[4]\colon&\fm{(\DIA A\IMP \BOX B)\IMP\BOX(A\IMP B)}\\%x[1ex]
	\end{array}
	\end{equation*}
\end{itemize}

The relational semantics for $\IK$ was first defined by Fischer-Servi~\cite{fischer1984}.
%
It combines the Kripke semantics for intuitionistic propositional logic and the one for classical modal logic, using two distinct relations on the set of worlds.

\begin{definition}
	A \emph{bi-relational frame} $\F$ is a triple $\langle W, R, \le \rangle$ 
	%	of a non-empty set of worlds $W$ equipped with two binary relations $R$ and $\le$, where $R$ being the modal \emph{accessibility relation} and $\le$ a preorder (\emph{i.e.} a reflexive and transitive relation), satisfying the following conditions:
	of a set of worlds $W$ equipped with an {accessibility relation} $\rel$ and a preorder $\le$ satisfying:
	\begin{enumerate}
		\item[($F_1$)] For $\lb u, \lb v, \lb{v'} \in W$, if $\accs uv$ and $\futs{v}{v'}$, there exists $\lb{u'}$ s.t.~$\futs{u}{u'}$ and $\accs{u'}{v'}$.

		\item[($F_2$)] For $\lb{u'}, \lb u, \lb v \in W$, if $\futs uv$, there exists $\lb{v'}$ s.t.~$\accs{u'}{v'}$ and $\futs{v}{v'}$.
	\end{enumerate}
	%	
\end{definition}

%\marianela{We recall the birelational models \cite{plotkin1986, ewald1986} for intuitionistic modal logics, which are a combination of the Kripke semantics for propositional intuitionistic logic and the one for classical modal logic. A \emph{birelational frame} $\F = \langle W, R, \le \rangle$ is a non-empty set $W$ of worlds together with two binary relations $\le, \rel \subseteq W \times W$, where $\le$ is a preorder(i.e., reflexive and transitive), such that the following two conditions hold:}

\begin{definition}
	A \emph{bi-relational model} $\M$ is a quadruple $\langle W, R,\le,V \rangle$ with $\langle W, R, \le \rangle$ a bi-relational frame and $V\colon W \to 2^\mathcal{A}$ a monotone valuation function, that is, a function mapping each world $\lb w$ to the subset of propositional atoms true at $\lb w$, additionally subject to:
	if $\futs{w}{w'}$ then $V(\lb w)\subseteq V(\lb{w'})$.
\end{definition}

We write $\lb w \Vdash \fm a$ if $\fm a \in V(\lb w)$, and by definition, we always have $\lb w \Vdash \fm{\TOP}$ and never that $\lb w \Vdash \fm{\BOT}$. 
%
Then the relation is extended to all formulas by induction, following the rules for both intuitionistic and modal Kripke models:

$\lb w \Vdash \fm{A \AND B}$ iff $\lb w \Vdash \fm A$ and $\lb w \Vdash \fm B$

$\lb w \Vdash \fm{A \OR B}$ iff $\lb w \Vdash \fm A$ or $\lb w \Vdash \fm B$

$\lb w \Vdash \fm{A \IMP B}$ iff for all $\lb{w'}$ with $\futs{w}{w'}$, if $\lb{w'} \Vdash \fm A$ then $\lb{w'} \Vdash \fm B$

$\lb w \Vdash \fm{\BOX A}$ iff for all $\lb{w'}$ and $\lb u$ with $\futs{w}{w'}$ and $\accs{w'}{u}$, $\lb u \Vdash \fm A$ %\hfill $(\ast)$

$\lb w \Vdash \fm{\DIA A}$ iff there exists a $\lb u$ such that $\accs wu$ and $\lb u \Vdash \fm A$.

\begin{definition}
	A formula $\fm A$ is \emph{satisfied} in a model $\M = \langle W, R, \le, V \rangle$, if for all $\lb w \in W$ we have $\lb w \Vdash \fm A$.
	%
	A formula $\fm A$ is \emph{valid} in a frame $\F = \langle W, R, \le \rangle$, if for all valuations $V$, $\fm A$ is satisfied in $\langle W, R, \le, V \rangle$.
\end{definition}

@@

\begin{theorem}[Fischer-Servi~\cite{fischer1984}, Plotkin and Stirling~\cite{plotkin1986}]\label{thm:plotkin}
	A formula $\fm A$ is a theorem of $\IK$ if and only if $\fm A$ is valid in every bi-relational frame.
\end{theorem}

\section{Labelled sequent calculi with $\rel$ and $\le$}

\begin{figure}%[h]
	\centering
	\small
	\fbox{
		\begin{minipage}{.95\textwidth}
			\begin{tabular}{@{\!}c@{\quad}c}
				$\vlinf{\rn{id}}{}{\B, \Left, \labels{x}{A} \SEQ \Right, \labels{x}{A} }{}$
				&
				$\vlinf{\llabrn\bot}{}{\B, \Left, \labels{x}{\BOT} \SEQ \Right}{}$
%				\quad
%				$\vlinf{\rlabrn\top}{}{\B, \Left \SEQ \Right, \labels{x}{\TOP}}{}$
				\\\\
				$\vlinf{\llabrn\AND}{}{\B,\Left, \labels{x}{A \AND B} \SEQ \Right}{\B, \Left, \labels{x}{A}, \labels{x}{B} \SEQ \Right}$
				&
				$\vliinf{\rlabrn\AND}{}{\B,\Left \SEQ \Right, \labels{x}{A \AND B}}{\B, \Left \SEQ \Right, \labels{x}{A}}{\B, \Left \SEQ \Right, \labels{x}{B}}$
				\\\\
				$\vliinf{\llabrn\OR}{}{\B, \Left, \labels{x}{A \OR B} \SEQ \Right}{\B, \Left, \labels{x}{A} \SEQ \Right}{\B, \Left, \labels{x}{B} \SEQ \Right}$
				&
				$\vlinf{\rlabrn\OR}{}{\B, \Left \SEQ \Right, \labels{x}{A \OR B}}{\B, \Left \SEQ \Right, \labels{x}{A}, \labels{x}{B}}$
				\\\\
				\multicolumn{2}{c}{
					$\vlinf{\llabrn\IMP}{\text{\scriptsize $\lb y$ fresh}}{\B, \Left \SEQ \Right, \labels{x}{A \IMP B}}{\B, \Left, \futs xy, \labels{y}{A} \SEQ \Right, \labels{y}{B}}$
				}
				\\\\
				\multicolumn{2}{c}{
					$\vliinf{\rlabrn\IMP}{}{\B, \futs xy, \Left, \labels{x}{A} \SEQ B \SEQ \Right}{\B, \futs xy, \Left \SEQ \Right, \labels{y}{A}}{\B, \futs xy, \Left, \labels{y}{B} \SEQ \Right}$ 
					% 		&$\vliinf{\rlabrn\IMP}{\text{\scriptsize $x \le y \in \B$}}{\B, \Left, \labels{x}{A \IMP B} \SEQ \Right}{\B, \Left \SEQ \Right, \labels{y}{A}}{\B, \Left, \labels{y}{B} \SEQ \Right}$
				}
				\\\\
				$\vlinf{\llabrn\BOX}{}{\B, \futs xy, \accs yz, \Left, \labels{x}{\BOX A} \SEQ \Right}{\B, \futs xy, \accs yz, \Left, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right}$
				&
				$\vlinf{\rlabrn\BOX}{\text{\scriptsize $\lb y, \lb z$ fresh}}{\B, \Left \SEQ \Right, \labels{x}{\BOX A}}{\B, \futs xy, \accs yz, \Left \SEQ \Right, \labels{z}{A}}$
				\\\\
				$\vlinf{\llabrn\DIA}{\text{\scriptsize $\lb y$ fresh}}{\B, \Left, \labels{x}{\DIA A} \SEQ \Right}{\B, \accs xy, \Left, \labels{y}{A} \SEQ \Right}$
				&
				$\vlinf{\rlabrn\DIA}{}{\B, \accs xy, \Left \SEQ \Right, \labels{x}{\DIA A}}{\B, \accs xy, \Left \SEQ \Right, \labels{x}{\DIA A}, \labels{y}{A}}$
				\\
				\multicolumn{2}{c}{
					$\mbox{\hbox to .9\linewidth{\dotfill}}$
				}
				\\
				$\vlinf{\rn{refl}}{}{\B, \Left \SEQ \Right}{\B, \futs xx, \Left \SEQ \Right}$
				&
				$\vlinf{\rn{trans}}{}{\B, \futs xy, \futs yz, \Left \SEQ \Right}{\B, \futs xy, \futs yz, \futs xz, \Left \SEQ \Right}$
				\\\\
				\multicolumn{2}{c}{
					$\vlinf{\rn{F_1}}{\text{\scriptsize $\lb u$ fresh}}{\B, \accs xy, \futs yz, \Left \SEQ \Right}{\B, \accs xy, \futs yz, \futs xu, \accs uz, \Left \SEQ \Right}$
				}
				\\\\
				\multicolumn{2}{c}{
					$\vlinf{\rn{F_2}}{\text{\scriptsize $\lb u$ fresh}}{\B, \accs xy,\futs xz, \Left \SEQ \Right}{\B, \accs xy, \futs xz, \futs yu, \accs zu, \Left \SEQ \Right }$		
				}
			\end{tabular}		
		\end{minipage}
	}		
	\caption{System $\labIKp$}
	\label{fig:labIKp}
\end{figure}

@@

\emph{Labelled deduction} has been proposed by Gabbay~\cite{gabbay1996} in the 80s as a unifying framework throughout proof theory in order to provide proof
systems for a wide range of logics. 
%
For modal logics it can take
the form of labelled natural deduction and labelled sequent systems as used, for example, by Simpson~\cite{simpson1994}, Vigan\`o~\cite{vigano2000}, and Negri~\cite{negri2005}. 
%
Simpson's proposed labelled sequent system for intuitionistic modal logic is representing explicitly only the accessibility relation $\rel$ in the syntax.
%
Echoing the definition of bi-relational structures, we consider another extension of labelled deduction to the intuitionistic setting. 
%
The idea is to use two sorts of relational atoms, one for the modal relation $\rel$ and another one for the intuitionistic relation $\leq$ as proposed by Maffezioli, Naibo and Negri in~\cite{maffezioli2013}. 
%

\begin{definition}
	A two-sided intuitionistic \emph{labelled sequent} is of the form $\B, \Left \SEQ \Right$ where $\B$ denotes a set of relational atoms $\accs xy$ and preorder atoms $\futs xy$, and $\Left$ and $\Right$ are multi-sets of labelled formulas $\labels{x}{A}$ (for $\lb x$ and $\lb y$ taken from the set of labels and $A$ an intuitionistic modal formula).
\end{definition}

@@

%
We obtain a proof system $\labIKp$, displayed on Figure~\ref{fig:labIKp}, for intuitionistic modal logic in this formalism. 
%
Most rules are similar to the ones of Simpson~\cite{simpson1994}, but some rules are even more explicitly in correspondence with the semantics by using the preorder atoms. 
%
In particular, the rules introducing the $\BOX$-operator correspond to the definition $(\ast)$.
%
Furthermore, our system has to incorporate the two semantic conditions ($F_1$) and ($F_2$) into the deductive rules $\rn{F_1}$ and $\rn{F_2}$, and the rules $\rn{refl}$ and $\rn{trans}$ are also necessary to ensure that the preorder atoms do behave as a preorder relation on labels.
%
%As we mentioned, we obtain a proof system $\labIKp$ which allows us to give an extension of labelled deduction to the intuitionistic world and then we prove the next theorem:

@@

\begin{theorem}\label{thm:cutfree-compl}
	%	Let $\CC$ be a set of geometric frame properties as in~\eqref{eq:cla-geometric} and $\labbrn{\CC}$ be the corresponding set of rules following schema~\eqref{eq:modal-grs}.
	%
	For any formula $A$, the following are equivalent.
	%
	\begin{enumerate}
		\item\label{i} $A$ is a theorem of $\IK$ 
		%
		\item\label{ii} $A$ is provable in $\labIKp +\labrn{cut}$ with %\quad
		%		
		%		\todo{}
		\smash{%\fbox{
			\scalebox{.9}{$\vliinf{\labrn{cut}}{}{\B_1, \B_2, \Left \SEQ \Right}{\B_1, \Left \SEQ \Right, \labels{z}{C}}{\B_2, \Left, \labels{z}{C} \SEQ \Right}$}
		}%}
		%
		\item\label{iii} $A$ is provable in $\labIKp$
		%
		\item\label{iv} $A$ is valid in every birelational frames %satisfying the properties in $\CC$.
	\end{enumerate}
\end{theorem}

%We show that each rule from our system is sound.
%
%We also present a syntactic completeness proof with respect the Hilbert system: we prove all Hilbert axioms using the rules from our system (i.e. proof of all propositional intuitionistic axioms, the five variants of k axiom from the intuitionistic syntax, simulate the necessitation rule and simulate modus ponens).
%
%We present a completeness proof for our system $\labIKp$ using the Simpson system. 
%
%The idea comes from knowing that the Simpson system is a Cut-free system, so this proof lets us know that our system is complete without the cut rule. 
%
%We show the proof by case analysis. 
%
%Most of the rules from Simpson system are the same as the rules in the system $\labIKp$, then we prove for the rules that are different.

>>>>

%\usepackage[matrix,arrow]{xy}
\usepackage[noxy]{../../virginialake}
%\vlnosmallleftlabels
\vlnostructuressyntax

@@

\begin{frontmatter}
  \title{Fully structured proof theory for intuitionistic modal logics}
  \author{Sonia Marin}
  %\footnote{You can put your email address or grant acknowledgement as a footnote, if you wish.}
  \address{University College London, UK}
  \author{Marianela Morales}
  \address{LIX, \'Ecole Polytechnique  \&  Inria Saclay, France}

@@

Structural proof theoretic accounts of modal logic can adopt the paradigm of \emph{labelled deduction} in the form of e.g.~labelled sequent systems~\cite{vigano2000,negri2005}, or the one of \emph{unlabelled deduction} in the form of e.g.~nested sequent systems~\cite{brunnler2009deep,poggiolesi2009method}.

These generalisations of the sequent framework, inspired by relational semantics, are needed to treat modalities uniformly. 
%
By extending the ordinary sequent structure with \emph{one} extra element, either relational atoms between labels or nested bracketing, they encode respectively graphs or trees in the sequents, giving them enough power to represent modalities.

Similarly, proof systems have been designed for intuitionistic modal logic both as labelled sequent systems~\cite{simpson1994} and as nested sequent systems~\cite{strassburger2013,kuznets:strassburger:maehara}.
%
Surprisingly, in both settings, extending the sequent structure with the same \emph{one} extra element is enough to obtain sound and complete systems.

This no longer matches the relational semantics of these logics, which requires to combine the relation for intuitionistic propositional logic and the one for modal logic. 
%
More importantly, it leads to deductive systems that are not entirely satisfactory; they cannot as modularly capture axiomatic extensions (or equivalently restricted semantical conditions) and, in particular, can only provide decision procedures for a handful of them~\cite{simpson1994}.

This lead us to develop a \emph{fully structured} approach to intuitionistic modal proof theory capturing both the modal accesibility relation and the intuitionistic pre-order relation. 
%
A fully labelled framework, described succintly in Section~\ref{sec:labelled}, has already allowed us to obtain modular systems for all intuitionistic Scott-Lemmon logics~\cite{marin:morales:strassburger:hal}. 
%
In an attempt to make this system amenable for proof-search and decision procedures, we have started investigated a fully nested framework, presented in Section~\ref{sec:nested}.

%We present a labelled sequent system and a nested sequent system for intuitionistic modal logics capturing both the modal accesibility relation and the intuitionistic pre-order relation. 
%
%In the labelled calculus we made explicity both relations with $\rel$ and $\le$ into the system. 
%
%In the nested sequent system, we represent this relations with one bracket in a sequent (for the accesibility relation) and a with two brackets in a sequent (preorder relation).  
%

\section{Intuitionistic modal logic}
The language of {intuitionisitic modal logic} is the one of intuitionistic propositional logic with the modal operators $\BOX$ and $\DIA$. %standing most generally for \emph{necessity} and \emph{possibility}.
%
Starting with a set $\mathcal{A}$ of atomic propositions, denoted $\fm a$, modal formulas are constructed from the grammar:
%
$$
\fm A \coloncolonequals
\fm a \mid \fm{\BOT} \mid \fm{(A \AND A)} \mid \fm{(A \OR A)} \mid \fm{(A \IMP A)} \mid \fm{\BOX A} \mid \fm{\DIA A}
$$
%
The axiomatisation of intuitionistic modal logic $\IK$ \cite{plotkin1986,fischer1984}
%, and by Ewald~\cite{Ewald} in the case of intuitionistic tense logic. 
%
is obtained from intuitionistic propositional logic by adding:
\begin{itemize}
	\item the \emph{necessitation rule}: $\fm{\BOX A}$ is a theorem if $\fm{A}$ is a theorem; and
	\item the following five variants of the \emph{distributivity axiom}:
	\begin{equation*}
	\label{eq:ik}%\hskip-2em
	\begin{array}[t]{r@{\;}l@{\quad}r@{\;}l@{\quad}r@{\;}l}
	\kax[1]\colon&\fm{\BOX(A\IMP B)\IMP(\BOX A\IMP\BOX B)}
	&
	\kax[3]\colon&\fm{\DIA(A\OR B)\IMP(\DIA A\OR\DIA B)}
	&
	\kax[5]\colon&\fm{\DIA\BOT\IMP\BOT}
	\\
	\kax[2]\colon&\fm{\BOX(A\IMP B)\IMP(\DIA A\IMP\DIA B)}
	&
	\kax[4]\colon&\fm{(\DIA A\IMP \BOX B)\IMP\BOX(A\IMP B)}\\%x[1ex]
	\end{array}
	\end{equation*}
\end{itemize}

The relational semantics for $\IK$~\cite{fischer1984} combines the Kripke semantics for intuitionistic propositional logic and the one for classical modal logic. % using two distinct relations on the set of worlds.

\begin{definition}
	A \emph{bi-relational frame} $\F$ consists %is a triple $\langle W, R, \le \rangle$ 
	%	of a non-empty set of worlds $W$ equipped with two binary relations $R$ and $\le$, where $R$ being the modal \emph{accessibility relation} and $\le$ a preorder (\emph{i.e.} a reflexive and transitive relation), satisfying the following conditions:
	of a set of worlds $W$ equipped with an {accessibility relation} $\rel$ and a preorder $\le$ satisfying:
	\begin{enumerate}
		\item[($F_1$)] For $\lb x, \lb y, \lb z \in W$, if $\accs xy$ and $\futs yz$, there exists $\lb u$ s.t.~$\futs xu$ and $\accs uz$.

		\item[($F_2$)] For $\lb x, \lb y, \lb z \in W$, if $\futs xy$ and $\accs xz$, there exists $\lb u$ s.t.~$\accs yu$ and $\futs zu$.
	\end{enumerate}
	%	
\end{definition}

\begin{definition}
	A \emph{bi-relational model} $\M$ is %a quadruple $\langle W, R,\le,V \rangle$ with $\langle W, R, \le \rangle$ 
	a bi-relational frame and a monotone valuation function $V\colon W \to 2^\mathcal{A}$.%, that is, a function mapping each world $\lb w$ to the subset of propositional atoms true at $\lb w$, additionally subject to: if $\futs{w}{w'}$ then $V(\lb w)\subseteq V(\lb{w'})$.
\end{definition}

We write $\lb x \Vdash \fm a$ if $\fm a \in V(\lb x)$ and, by definition, it is never the case that $\lb x \Vdash \fm{\BOT}$. %and always have $\lb w \Vdash \fm{\TOP}$
%
The relation $\Vdash$ is extended to all formulas by induction, following the rules for both intuitionistic and modal Kripke models:
\begin{align}
	&\lb x \Vdash \fm{A \AND B} && \text{iff}\qquad 
	\lb x \Vdash \fm A \text{ and } \lb x \Vdash \fm B
	\nonumber\\
	&\lb x \Vdash \fm{A \OR B} && \text{iff}\qquad 
	\lb x \Vdash \fm A \text{ or } \lb x \Vdash \fm B
	\nonumber\\
	&\lb x \Vdash \fm{A \IMP B} && \text{iff}\qquad 
	\text{for all } \lb y \text{ with } \futs xy\text{, if }\lb y \Vdash \fm A\text{ then }\lb y \Vdash \fm B
	\nonumber\\
	&\lb x \Vdash \fm{\BOX A} && \text{iff}\qquad
	\text{for all }\lb y\text{ and }\lb z\text{ with }\futs xy\text{ and }\accs yz, \lb z \Vdash \fm A %\hfill $(\ast)$
	\tag{$\star$}\label{eq:box}\\
	&\lb x \Vdash \fm{\DIA A} && \text{iff}\qquad
	\text{there exists a }\lb y\text{ such that }\accs xy\text{ and }\lb y \Vdash \fm A
	\nonumber
\end{align}

\begin{definition}
%	A formula $\fm A$ is \emph{satisfied} in a model $\langle W, R, \le, V \rangle$, if for all $\lb w \in W$ we have $\lb w \Vdash \fm A$.
	%
	A formula $\fm A$ is \emph{valid} in a frame $\langle W, R, \le \rangle$, if for all monotone valuations $V$ and for all $\lb w \in W$, we have $\lb w \Vdash \fm A$
\end{definition}

@@

\begin{theorem}[\cite{fischer1984,plotkin1986}]\label{thm:plotkin}
	A formula $\fm A$ is a theorem of $\IK$ if and only if $\fm A$ is valid in every bi-relational frame.
\end{theorem}

\section{Fully labelled sequent calculus}\label{sec:labelled}

\begin{figure}%[h]
	\centering
	\small
	\fbox{
		\begin{minipage}{.95\textwidth}
			\begin{tabular}{@{\!}c@{\quad}c}
				$\vlinf{\rn{id}}{}{\B, \futs xy, \Left, \labels{x}{A} \SEQ \Right, \labels{y}{A} }{}$
				&
				$\vlinf{\llabrn\bot}{}{\B, \Left, \labels{x}{\BOT} \SEQ \Right}{}$
%				\quad
%				$\vlinf{\rlabrn\top}{}{\B, \Left \SEQ \Right, \labels{x}{\TOP}}{}$
				\\\\
				$\vlinf{\llabrn\AND}{}{\B,\Left, \labels{x}{A \AND B} \SEQ \Right}{\B, \Left, \labels{x}{A}, \labels{x}{B} \SEQ \Right}$
				&
				$\vliinf{\rlabrn\AND}{}{\B,\Left \SEQ \Right, \labels{x}{A \AND B}}{\B, \Left \SEQ \Right, \labels{x}{A}}{\B, \Left \SEQ \Right, \labels{x}{B}}$
				\\\\
				$\vliinf{\llabrn\OR}{}{\B, \Left, \labels{x}{A \OR B} \SEQ \Right}{\B, \Left, \labels{x}{A} \SEQ \Right}{\B, \Left, \labels{x}{B} \SEQ \Right}$
				&
				$\vlinf{\rlabrn\OR}{}{\B, \Left \SEQ \Right, \labels{x}{A \OR B}}{\B, \Left \SEQ \Right, \labels{x}{A}, \labels{x}{B}}$
				\\\\
				\multicolumn{2}{c}{
					$\vlinf{\llabrn\IMP}{\text{\scriptsize $\lb y$ fresh}}{\B, \Left \SEQ \Right, \labels{x}{A \IMP B}}{\B, \Left, \futs xy, \labels{y}{A} \SEQ \Right, \labels{y}{B}}$
				}
				\\\\
				\multicolumn{2}{c}{
					$\vliinf{\rlabrn\IMP}{}{\B, \futs xy, \Left, \labels{x}{A} \SEQ B \SEQ \Right}{\B, \futs xy, \Left \SEQ \Right, \labels{y}{A}}{\B, \futs xy, \Left, \labels{y}{B} \SEQ \Right}$ 
					% 		&$\vliinf{\rlabrn\IMP}{\text{\scriptsize $x \le y \in \B$}}{\B, \Left, \labels{x}{A \IMP B} \SEQ \Right}{\B, \Left \SEQ \Right, \labels{y}{A}}{\B, \Left, \labels{y}{B} \SEQ \Right}$
				}
				\\\\
				$\vlinf{\llabrn\BOX}{}{\B, \futs xy, \accs yz, \Left, \labels{x}{\BOX A} \SEQ \Right}{\B, \futs xy, \accs yz, \Left, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right}$
				&
				$\vlinf{\rlabrn\BOX}{\text{\scriptsize $\lb y, \lb z$ fresh}}{\B, \Left \SEQ \Right, \labels{x}{\BOX A}}{\B, \futs xy, \accs yz, \Left \SEQ \Right, \labels{z}{A}}$
				\\\\
				$\vlinf{\llabrn\DIA}{\text{\scriptsize $\lb y$ fresh}}{\B, \Left, \labels{x}{\DIA A} \SEQ \Right}{\B, \accs xy, \Left, \labels{y}{A} \SEQ \Right}$
				&
				$\vlinf{\rlabrn\DIA}{}{\B, \accs xy, \Left \SEQ \Right, \labels{x}{\DIA A}}{\B, \accs xy, \Left \SEQ \Right, \labels{x}{\DIA A}, \labels{y}{A}}$
				\\
				\multicolumn{2}{c}{
					$\mbox{\hbox to .9\linewidth{\dotfill}}$
				}
				\\
				$\vlinf{\rn{refl}}{}{\B, \Left \SEQ \Right}{\B, \futs xx, \Left \SEQ \Right}$
				&
				$\vlinf{\rn{trans}}{}{\B, \futs xy, \futs yz, \Left \SEQ \Right}{\B, \futs xy, \futs yz, \futs xz, \Left \SEQ \Right}$
				\\\\
				\multicolumn{2}{c}{
					$\vlinf{\rn{F_1}}{\text{\scriptsize $\lb u$ fresh}}{\B, \accs xy, \futs yz, \Left \SEQ \Right}{\B, \accs xy, \futs yz, \futs xu, \accs uz, \Left \SEQ \Right}$
				}
				\\\\
				\multicolumn{2}{c}{
					$\vlinf{\rn{F_2}}{\text{\scriptsize $\lb u$ fresh}}{\B, \accs xy,\futs xz, \Left \SEQ \Right}{\B, \accs xy, \futs xz, \futs yu, \accs zu, \Left \SEQ \Right }$		
				}
			\end{tabular}		
		\end{minipage}
	}		
	\caption{System $\labIKp$}
	\label{fig:labIKp}
\end{figure}

@@

%\emph{Labelled deduction} has been proposed by Gabbay~\cite{gabbay1996} in the 80s as a unifying framework throughout proof theory in order to provide proof systems for a wide range of logics. 
%
%For modal logics it can take the form of labelled natural deduction and labelled sequent systems as used, for example, by Simpson~\cite{simpson1994}, Vigan\`o~\cite{vigano2000}, and Negri~\cite{negri2005}. 
%
Simpson's~\cite{simpson1994} proposed labelled sequent system for intuitionistic modal logic is representing explicitly only the accessibility relation $\rel$ in the syntax.
%
Echoing the definition of bi-relational structures, we consider another extension of labelled deduction to the intuitionistic setting. 
%
The idea is to use two sorts of relational atoms, one for the modal relation $\rel$ and another one for the intuitionistic relation $\leq$ (similarly to~\cite{maffezioli2013} for epistemic logic). 
%

\begin{definition}
	A two-sided intuitionistic \emph{fully labelled sequent} is of the form $\B, \Left \SEQ \Right$ where $\B$ denotes a set of relational atoms $\accs xy$ and preorder atoms $\futs xy$, and $\Left$ and $\Right$ are multi-sets of labelled formulas $\labels{x}{A}$ (for $\lb x$ and $\lb y$ taken from a countable set of labels and $A$ an intuitionistic modal formula).
\end{definition}

@@

%
We obtain a proof system $\labIKp$, displayed on Figure~\ref{fig:labIKp}, for intuitionistic modal logic in this formalism. 
%
Most rules are similar to the ones of Simpson~\cite{simpson1994}, but some rules are even more explicitly in correspondence with the semantics by using the preorder atoms. 
%
In particular, the rules introducing the $\BOX$-operator correspond to the definition~\eqref{eq:box}.
%
Furthermore, our system has to incorporate the two semantic conditions ($F_1$) and ($F_2$) into the deductive rules $\rn{F_1}$ and $\rn{F_2}$, and the rules $\rn{refl}$ and $\rn{trans}$ are also necessary to ensure that the preorder atoms do behave as a preorder relation on labels.
%
%As we mentioned, we obtain a proof system $\labIKp$ which allows us to give an extension of labelled deduction to the intuitionistic world and then we prove the next theorem:

@@

\begin{theorem}\label{thm:cutfree-compl}
	%	Let $\CC$ be a set of geometric frame properties as in~\eqref{eq:cla-geometric} and $\labbrn{\CC}$ be the corresponding set of rules following schema~\eqref{eq:modal-grs}.
	%
	For any formula $A$, the following are equivalent.
	%
	\begin{enumerate}
		\item\label{i} $A$ is a theorem of $\IK$ 
		%
		\item\label{ii} $A$ is provable in $\labIKp +\labrn{cut}$ with %\quad
		%		
		\smash{%\fbox{
			\scalebox{.9}{$\vliinf{\labrn{cut}}{}{\B_1, \B_2, \Left \SEQ \Right}{\B_1, \Left \SEQ \Right, \labels{z}{C}}{\B_2, \Left, \labels{z}{C} \SEQ \Right}$}
		}%}
		%
		\item\label{iii} $A$ is provable in $\labIKp$
		%
		\item\label{iv} $A$ is valid in every birelational frames %satisfying the properties in $\CC$.
	\end{enumerate}
\end{theorem}
